Consistent, and weak, ecosystem functional response across precipitation extremes in a sagebrush steppe

### 2017-08-25 manuscript/usses_water_ms.tex
writing the discussion

At a given site, the functional response of aboveground net primary
productivity (ANPP) to water availability (e.g., soil moisture) can be
characterized by fitting a model to historical observations of ANPP and
soil moisture. However, the fitted functional response may provide an
incomplete picture because future conditions are likely to be outside
the historical range of variability \citep{Smith2011}. For example,
historical trends may underestimate the potential for the soil
moisture-ANPP relationship to saturate if soil moisture is pushed far
beyond typical levels. Saturating relationships are actually common
\citep{Hsu2012, Gherardi2015a}, perhaps because other resources, like
nitrogen, become more limiting in wet years than dry years. Knowing the
curvature of the soil moisture-ANPP relationship at extreme
precipitation levels is critical for understanding how ecosystems will
respond to chronic alterations in water availability.

Another problem with relying on historical ecosystem functional
responses is that they are not static. Changes in species identities and
abundances can alter an ecosystem's functional response to water
availability because different species have different physiological
thresholds for producing biomass. \citet{Smith2009} introduced the
`Hierarchical Response Framework' (HRF) for understanding the interplay
of community composition and ecosystem functioning in response to
resource manipulations over time. In the near term, ecosystem
functioning such as ANPP will reflect the physiological responses of
individual species to the manipulated resource level. For example, ANPP
may decline under simulated drought because the initial community
consisted of drought-intolerant species \citep{Hoover2014}. Over longer
time spans, ecosystem functioning may recover as new species colonize or
initial species reorder in relative abundance. For example, ANPP may
initially decline, but eventually rise back to pre-treatment levels once
drought-tolerant species become more abundant and compensate for
drought-intolerant species \citep{Hoover2014}. It is also possible that
ecosystem functioning shifts to a new mean state, reflecting the suite
of species in the new community \citep{Knapp2012}.

Manipulating potentially limiting resources, like precipitation, offers
a route to understanding how ecosystems will respond to resource levels
that fall outside the historical range of variability
\citep{Avolio2015, Gherardi2015, Knapp2017}. Altering the amount of
precipitation over many years should provide insight into the time
scales at which water-limited ecosystems respond to chronic resource
alteration. Following the HRF, we propose four alternative predictions
for the effect of precipitation manipulation on the ecosystem functional
response to soil moisture, that is, the soil moisture-ANPP relationship
(Fig. 1). The four predictions are based on possible outcomes at the
community (e.g., community composition) and ecosystem (e.g., soil
moisture-ANPP regression) levels.

First, altered precipitation changes neither ecosystem functional
response nor community composition (Fig. 1, top left). In this case,
changes in ANPP simply follow the soil moisture-ANPP relationship under
ambient conditions. This corresponds to the early phases of the HRF,
where ecosystem response is due to the physiological responses of
individual species. Second, the ecosystem functional response changes
but community composition remains the same (Fig. 1, top right). A
saturating soil moisture-ANPP response fits this scenario, where
individual species hit physiological thresholds or are limited by some
other resource. Third, the ecosystem functional response is consistent
but underlying community composition changes (Fig. 1, bottom left). In
this case, changes in species' identities or abundances occur in
response to altered precipitation levels and species more suited to the
new conditions compensate for reduced function of initial species.
Fourth, and last, both ecosystem functional response and community
composition change (Fig. 1, bottom right). New species, or newly
abundant species, with different physiological responses completely
reshape the ecosystem functional response.

All four outcomes are possible in any given ecosystem, but the time
scales at which the different scenarios play out likely differ
\citep{Smith2009, Wilcox2016, Knapp2017}. Thus, our task is not to test
the validity of the HRF, but rather to amass information on how quickly
ecosystem functional responses change in different ecosystems. Likewise,
we need to understand whether changes at the ecosystem level are driven
by community level changes or individual level responses.

To that end, here we report the results of a five-year precipitation
manipulation experiment in a sagebrush steppe grassland. We imposed
drought and irrigation treatments (approximately \(\pm50\%\)) and
measured ecosystem (ANPP) and community (species composition) responses.
We focus on how the drought and irrigation treatments affect the
relationship between available soil moisture and ANPP, and if community
dynamics underlie the ecosystem responses. In particular, we are
interested in the consistency of the soil moisture-ANPP relationship
among treatments. Is the relationship steeper under the drought
treatment, at low soil moisture? Does the relationship saturate under
the irrigation treatment, at high soil moisture? To answer these
questions we fit a random intercept, random slope model to test whether
the regressions differed among treatments. We also analyzed community
composition over time, allowing us to place our experimental results
within the framework of the HRF and our competing predictions (Fig. 1).

@@

Drought and irrigation treatments were designed to decrease and increase
the amount of ambient precipitation by 50\%, respectively. To achieve
this, we used a system of rain-out shelters and automatic irrigation
\citep{Gherardi2013}. The rain-out shelters consisted of transparent
acrylic shingles 1-1.5 m above the ground that covered an area of
\(2.5\times2\) m. The shingles intercepted approximately 50\% of
incoming rainfall, which was channeled into 75 liter containers.
Captured rainfall was then pumped out of the containers and sprayed on
to the adjacent irrigation plot via two suspended sprinklers. Pumping
was triggered by float switches once water levels reached about 20
liters. We disconnected the irrigation pumps each fall and reconnected
them, often with difficulty, each spring. The rain-out shelters remained
in place throughout the year.

We monitored soil moisture in four of the drought-irrigation pairs using
Decagon Devices (Pullman, Washington) 5TM and EC-5 soil moisture
sensors. We installed four sensors around the edges of each plot, two at
5 cm soil depth and two at 25 cm soil depth. We also installed four
sensors in areas nearby the four selected plot pairs to measure ambient
soil moisture at the same depths. Soil moisture measurements were
automatically logged every four hours. We coupled this temporally
intensive soil moisture sampling with spatially extensive readings taken
at six points within all 16 plots and associated ambient measurement
areas. These snapshot data were collected on 06/06/2012, 04/29/2015,
05/07/2015, 06/09/2015, and 05/10/2016 using a handheld EC-5 sensor.

@@

Three of our five treatment years fell in years of below average
rainfall (Fig. 2A). Thus, those three years represent a lower magnitude
of absolute change in precipitation experienced by the treatments.
Averaged across treatments, ANPP varied from a minimum of 74.5 g
m\(^{-2}\) in 2014 to a maximum of 237.1 g m\(^{-2}\) in 2016 (Fig. 2C).
ANPP was slightly higher in irrigation plots and slightly lower in
drought plots (Fig. 2C), corresponding to estimated soil volumetric
water content (VWC) differences among treatments (Fig. 2B). Such
differences in soil VWC indicate our treatment infrastructure was
successful.

Cumulative March-June soil moisture had a weak positive effect on ANPP
(Table 1; Fig. 3), but the effect is associated with high uncertainty,
with credible intervals that overlap zero, regardless of treatment
(Table 1). Ecosystem functional response was similar among treatments
(Table 1; Fig. 3B), but there is evidence that the slope for the
irrigation treatment is less than the slope for the control treatment.
This evidence comes from interpreting the posterior distribution of the
slope offset for the irrigation treatment, from which we calculate a
98\% probability that the estimate is less than zero (Fig. 3A, right
panel). There was no evidence that the treatment effects became more
important over time because there was no directional trend in the random
year effects (Fig. A2-2).

>>>>

The functional response of aboveground net primary productivity (ANPP)
to water availability (e.g., soil moisture) can be characterized by
fitting a model to historical observations of ANPP and soil moisture at
a given site. However, the fitted functional response may provide an
incomplete picture because future conditions are likely to be outside
the historical range of variability \citep{Smith2011}. For example,
historical trends may underestimate the potential for the soil
moisture-ANPP relationship to saturate or weaken if soil moisture is
pushed far beyond typical levels. Saturating relationships are actually
common \citep{Hsu2012, Gherardi2015a}, perhaps because other resources,
like nitrogen, become more limiting in wet years than dry years. Knowing
the curvature of the soil moisture-ANPP relationship at extreme
precipitation levels is critical for understanding how ecosystems will
respond to chronic alterations in water availability.

Another problem with relying on historical ecosystem functional
responses is that they are not static. Changes in species identities and
abundances can alter an ecosystem's functional response to water
availability because different species have different physiological
thresholds for producing biomass. \citet{Smith2009} introduced the
`Hierarchical Response Framework' for understanding the interplay of
community composition and ecosystem functioning in response to resource
manipulations over time. In the near term, ecosystem functioning such as
ANPP will reflect the physiological responses of individual species to
the manipulated resource level. For example, ANPP may decline under
simulated drought because the initial community consisted of
drought-intolerant species \citep{Hoover2014}. Ecosystem functioning may
recover over longer time spans as new species colonize or initial
species reorder in relative abundance. For example, ANPP may initially
decline, but eventually rise back to pre-treatment levels once
drought-tolerant species become more abundant and compensate for
drought-intolerant species \citep{Hoover2014}. It is also possible that
ecosystem functioning shifts to a new mean state, reflecting the suite
of species in the new community \citep{Knapp2012}.

Manipulating potentially limiting resources, like precipitation, offers
a route to understanding how ecosystems will respond to resource levels
that fall outside the historical range of variability
\citep{Avolio2015, Gherardi2015, Knapp2017}. Altering the amount of
precipitation over many years should provide insight into the time
scales at which water-limited ecosystems respond to chronic resource
alteration. We propose four alternative predictions for the effect of
precipitation manipulation on the ecosystem functional response to soil
moisture based on the Hierarchical Response Framework (Fig. 1). We
define `ecosystem functional response' as the relationship between
available soil moisture and ANPP. The four predictions are based on
possible outcomes at the community (e.g., community composition) and
ecosystem (e.g., soil moisture-ANPP regression) levels.

First, altered precipitation changes neither ecosystem functional
response nor community composition (Fig. 1, top left). In this case,
changes in ANPP simply follow the soil moisture-ANPP relationship under
ambient conditions. This corresponds to the early phases of the
Hierarchical Respone Framework, where ecosystem response follows the
physiological responses of individual species. Second, the ecosystem
functional response changes but community composition remains the same
(Fig. 1, top right). A saturating soil moisture-ANPP response fits this
scenario, where individual species hit physiological thresholds or are
limited by some other resource. Third, the ecosystem functional response
is consistent but underlying community composition changes (Fig. 1,
bottom left). In this case, changes in species' identities or abundances
occur in response to altered precipitation levels and species more
suited to the new conditions compensate for reduced function of initial
species. Fourth, and last, both ecosystem functional response and
community composition change (Fig. 1, bottom right). New species, or
newly abundant species, with different physiological responses
completely reshape the ecosystem functional response.

All four outcomes are possible in any given ecosystem, but the time
scales at which the different scenarios play out likely differ
\citep{Smith2009, Wilcox2016, Knapp2017}. Thus, our task is not to test
the validity of the Hierarchical Response Framework, but rather to amass
information on how quickly ecosystem functional responses change in
different ecosystems. We also need to understand whether changes at the
ecosystem level are driven by community level changes or individual
level responses.

To that end, here we report the results of a five-year precipitation
manipulation experiment in a sagebrush steppe grassland. We imposed
drought and irrigation treatments (approximately \(\pm50\%\)) and
measured ecosystem (ANPP) and community (species composition) responses.
We focus on how the drought and irrigation treatments affect the
relationship between available soil moisture and ANPP, and if community
dynamics underlie the ecosystem responses. In particular, we are
interested in the consistency of the soil moisture-ANPP relationship
among treatments. Is the relationship steeper under the drought
treatment, at low soil moisture? Does the relationship saturate under
the irrigation treatment, at high soil moisture? To answer these
questions we fit a generalized linear mixed effects model to test
whether the regressions differed among treatments. We also analyzed
community composition over time, allowing us to place our experimental
results within the framework our competing predictions (Fig. 1).

@@

Drought and irrigation treatments were designed to decrease and increase
the amount of ambient precipitation by 50\%. To achieve this, we used a
system of rain-out shelters and automatic irrigation
\citep{Gherardi2013}. The rain-out shelters consisted of transparent
acrylic shingles 1-1.5 m above the ground that covered an area of
\(2.5\times2\) m. The shingles intercepted approximately 50\% of
incoming rainfall, which was channeled into 75 liter containers.
Captured rainfall was then pumped out of the containers and sprayed on
to the adjacent irrigation plot via two suspended sprinklers. Pumping
was triggered by float switches once water levels reached about 20
liters. We disconnected the irrigation pumps each fall and reconnected
them, often with difficulty, each spring. The rain-out shelters remained
in place throughout the year.

We monitored soil moisture in four of the drought-irrigation pairs using
Decagon Devices (Pullman, Washington) 5TM and EC-5 soil moisture
sensors. We installed four sensors around the edges of each plot, two at
5 cm soil depth and two at 25 cm soil depth. We also installed four
sensors in areas nearby the four selected plot pairs to measure ambient
soil moisture at the same depths. Soil moisture measurements were
automatically logged every four hours. We coupled this temporally
intensive soil moisture sampling with spatially extensive readings taken
at six points within all 16 plots and associated ambient measurement
areas. These snapshot data were collected on 06-06-2012, 04-29-2015,
05-07-2015, 06-09-2015, and
05-10-2016\footnote{Dates formatted as: mm-dd-yyyy.} using a handheld
EC-5 sensor.

@@

Three of our five treatment years fell in years of below average
rainfall (Fig. 2A). Thus, those three years represent a lower magnitude
of absolute change in precipitation experienced by the treatments. ANPP
varied from a minimum of 74.5 g m\(^{-2}\) in 2014 to a maximum of 237.1
g m\(^{-2}\) in 2016 when averaged across treatments (Fig. 2C). ANPP was
slightly higher in irrigation plots and slightly lower in drought plots
(Fig. 2C), corresponding to estimated soil volumetric water content
(VWC) differences among treatments (Fig. 2B). Such differences in soil
VWC indicate our treatment infrastructure was successful.

Cumulative March-June soil moisture had a weak positive effect on ANPP
(Table 1; Fig. 3). The effect of soil moisture for each treatment is
associated with high uncertainty, however, with credible intervals that
overlap zero (Table 1). Although the parameter estimates for the effect
of soil moisture overlap zero, the posterior distributions of the slopes
all shrank and shifted to more positive values relative to the prior
distributions (Fig. A2-2). Ecosystem functional response was similar
among treatments (Table 1; Fig. 3B), but there is evidence that the
slope for the irrigation treatment is less than the slope for the
control treatment. This evidence comes from interpreting the posterior
distribution of the slope offset for the irrigation treatment, from
which we calculate a 99\% one-tailed probability that the estimate is
less than zero (Fig. 3A, right panel). There was no evidence that the
treatment effects became more important over time because there was no
directional trend in the random year effects (Fig. A2-2).
